Automatically proving equivalence by type-safe reflection

### 2016-01-19 paper-JFP-DTP/content/1-introduction.tex
corrections with Edwin advices on latex

\subsection{Propositional and Judgemental Equalities in Intentional Type Theories}
In Mathematics, equality is a proposition, which can hold or otherwise, or be disproved.
Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
Thus, if $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition "$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$.  Thus, $Id$ is a type
family (parametrised by the type $a$) indexed over two elements of $a$, giving
$Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $Id_a\ x\ y$ as $x\ =_a\ y$.  This equality is the
equality which can be manipulated in the language.

@@

The judgemental equality has to be included in the propositional equal, because
what is equal by definition should be provably equal.  This is accomplished by
giving constructors to the type $Id(a,a)$ and nothing when "$a$ is not $b$".
One way to implement $Id$ in Idris is:

@@

Induction principles work for any proposition, so it also works with the
propositional equality. That means that the axiom of induction principle has
made the type $Id(a,b)$ bigger than it looks like (in the definition 1), and
there are therefore things which are provably equal which unfortunately aren't
equal syntactically after an expanding of all the definitions. The activity of
proving equalities is therefore in these theories something which isn't
automatically decidable by the type-checker in the general case.

@@

It should be pointed that the proof component of this function is established without any effort, as we always give the "Refl" for the proof. Indeed, for each of the eight patterns, a simple computation reduces the given and expected values to the same value, and "Refl" can therefore be used. We also note that
there is no need to produce any lemma of correctness about $addBit$ afterwards.
The correct by construction style in which it is written already gives the
property that we want: the two bits produced effectively represent the
addition of the three bits given in input.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) 
                 bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\figrule
\end{figure}

>>>>

\subsection{Propositional and Judgemental Equalities in Intentional Type Theories}
In Mathematics, equality is a proposition, which can hold or otherwise, or be disproved.
Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
Thus, if $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$.  Thus, $Id$ is a type
family (parametrised by the type $a$) indexed over two elements of $a$, giving
$Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $Id_a\ x\ y$ as $x\ =_a\ y$.  This equality is the
equality which can be manipulated in the language.

@@

The judgemental equality has to be included in the propositional equal, because
what is equal by definition should be provably equal.  This is accomplished by
giving constructors to the type $Id(a,a)$ and nothing when ``$a$ is not $b$".
One way to implement $Id$ in Idris is:

@@

Induction principles work for any proposition, so it also works with the
propositional equality. That means that the axiom of induction principle has
made the type $Id(a,b)$ bigger than it looks like in the above definition, and
there are therefore things which are provably equal which unfortunately aren't
equal syntactically after an expanding of all the definitions. The activity of
proving equalities is therefore in these theories something which isn't
automatically decidable by the type-checker in the general case.

@@

It should be pointed that the proof component of this function is established without any effort, as we always give the ``Refl" for the proof. Indeed, for each of the eight patterns, a simple computation reduces the given and expected values to the same value, and ``Refl" can therefore be used. We also note that
there is no need to produce any lemma of correctness about $addBit$ afterwards.
The correct by construction style in which it is written already gives the
property that we want: the two bits produced effectively represent the
addition of the three bits given in input.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) 
                 bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\label{adc_lemma_2}
\figrule
\end{figure}
